---
title: "Wisconsin Senate District 26 Results"
author: "Harald Kliems"
date: "8/12/2020"
output: html_document
---





Wisconsin had primary elections for the State Assembly and Senate on August 11. A good opportunity to practice creating some maps. I'll focus on the race for Senate District 26.

# Data
Results by ward are available from the Dane County Clerk's website; geographies for the wards are available on the City of Madison OpenData portal.

```{r}
library(tidyverse)
library(readxl)
library(tmap)
library(sf)
library(tidycensus)
get_acs(geography = "county", variable = "S0601_C01_016", state = "WI", geometry = TRUE)
```

```{r}
elect <- read_xlsx("data/ElectionResults.xlsx", 
                           skip = 7, 
                           col_names = c("precincts",
                                         "moe",
                                         "elm",
                                         "bur",
                                         "ben",
                                         "roy",
                                         "dav",
                                         "ime",
                                         "wri"))

wards <- read_sf("data/Wards.shp")

elect
wards %>% 
  filter(SEN_DIST == 26)
```
Data looks okay, but there are mismatches to deal with:

- There are wards with 0 total votes in `elect`
- The total number of wards between `elect` and `ward` doesn't match
- The ID numbers of wards in `elect` seem to be possibly not unique
- There is a ward with an ID of 1-2 in `elect`, and several wards in `wards` with an ID of `-1`

# Adding percentages

```{r}
elect <- elect %>% 
  mutate(total_votes = rowSums(.[2:9])) %>% 
  mutate(across(2:9, ~ .x/total_votes*100, .names = "{col}_pct"))
```


# Who won?
Answering the question who won in each district requires making the `elect` data long, then grouping by precinct, and then doing a comparison of the actual number of votes with the max number of votes in each precinct. Not exactly straightforward, but for a multi-candidate race, I can't think of a more elegant option.

```{r}


winners <- elect %>% 
  pivot_longer(2:9, names_to = "candidate", values_to = "votes") %>% 
  group_by(precincts) %>% 
  filter(votes != 0) %>% 
  filter(votes == max(votes)) %>% 
  select(-ends_with("pct")) %>% 
  mutate(winning_percent = round(votes/total_votes*100, 1))

winners

```

Each ward was one by either Kelda Roys or Nada Elmikasfi:

```{r}
unique(winners$candidate)
```

# Map it
Who won their districts?

```{r}
df <- winners %>% 
  mutate(ID = as.integer(str_extract(precincts, "[0-9]{3}"))) %>% #extracts the ward number
  mutate(ID = ifelse(precincts == "V Shorewood Hills Wds 1-2",
                     ))
  mutate(candidate = case_when(candidate == "elm" ~ "Nada Elmikashfi",
                               candidate == "roy" ~ "Kelda Roys"))

df2 <- wards %>% 
  inner_join(df, by = c("WARD" = "ID"))

tmap_mode("view")
tm_shape(df2) +
  tm_polygons("candidate", 
              legend.title = "Winning candidate",
              popup.vars = "winning_percent")


```
What was the winning margin of Kelda Roys over Nada Elmikashfi?
```{r}
elect2 <- elect %>% 
  mutate(margin = roy_pct - elm_pct,
         ID = as.integer(str_extract(precincts, "[0-9]{3}")))

df3 <- wards %>% 
  inner_join(elect2, by = c("WARD" = "ID"))

tm_shape(df3) +
  tm_polygons("margin", title = "Winning margin Kelda Roys")
```

